preciseTAD: A machine learning framework for precise TAD boundary prediction at base level resolution

### 2020-08-18 Manuscript/manuscript.Rmd
More polishing

**Results.** Spatial relationships (distance) between boundaries and annotations have the most predictive power. Transcription factor binding sites outperformed other genome annotation types. Random under-sampling significantly improved model performance. Boundaries predicted by *preciseTAD* were more enriched for CTCF, RAD21, SMC3, and ZNF143 signal and more conserved across cell lines, highlighting their higher biological significance. The model trained in one cell line performs well in predicting boundaries in another cell line.

@@

The advent of chromosome conformation capture (3C) sequencing technologies, and its successor Hi-C, have revealed a hierarchy of 3-dimensional (3D) spatial constructs within the human genome such as chromatin loops [@rao20143d], Topologically Associated Domains (TADs) [@nora2012spatial;@dixon2012topological], and A/B compartments [@lieberman2009comprehensive]. At the kilobase scale, gene promoters are often connected via distal regulatory elements, such as enhancers, through chromatin loops, thereby regulating gene expression [@Lupianez:2015aa; @Franke:2016aa]. Further, at the megabase scale, TADs represent regions on the linear genome, that are highly self-interacting. Emerging evidence has linked TADs to critical roles in cell dynamics and cell differentiation. Studies have shown that TADs themselves are highly conserved across species and cell lines [@dixon2012topological; @krefting2018evolutionary; @nora2012spatial; @dixon2015chromatin; @rao20143d; @Schmitt2016aa]. Likewise, disruption of boundaries demarcating TADs and loops promotes cancer [@hnisz2016activation; @taberlay2016three] and other disorders [@lupianez2016breaking;@Sun:2018aa;@Meaburn:2007aa]. Therefore, precisely identifying the location of TAD and chromatin loop boundaries remains a top priority in our goal to fully understand the functionality of the human genome.

Several methods have been proposed to identify genomic coordinates of TADs (reviewed in [@Zufferey:2018aa; @forcato2017comparison]), and chromatin loops [@rao20143d; @Ay:2014aa; @Heinz:2010aa; @salameh2019supervised].  However, a key limitation among most of these TAD- and loop callers is that they rely heavily on Hi-C data *resolution*. Resolution refers to the sizes of genomic regions (bins) used to segment the linear genome [@li2019comparison]. Lower resolutions contribute to larger bin sizes, which lead to uncertainty in the location of boundaries. Due to high sequencing costs, most available Hi-C datasets have relatively low resolution, ranging from 25 kb to 100 kb. Because increasing the resolution of Hi-C data requires a quadratic increase in the total sequencing depth, obtaining high resolution remains difficult [@schmitt2016genome]. While these low-resolution Hi-C datasets can be used to probe higher-order genomic patterns such as A/B compartments, more refined resolutions are necessary to accurately define TAD and loop boundaries [@fudenberg2016formation].

Another limitation among domain callers is that they disregard prior knowledge about functional genomic annotations associated with TAD and loop boundaries, referred hereafter as domain boundaries. The insulator binding protein, CTCF, and additional cofactors such as SMC3 and RAD21 have been identified as components of the loop extrusion model, whereby DNA is extruded through cohesin rings forming chromatin loops [@hansen2018recent;@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self; @Mirny:2019aa; @Davidson:2019aa]. Additionally, distinct patterns of histone modifications have also been shown to be present at boundaries, including H3K36me3 [@lieberman2009comprehensive;@dixon2012topological;@libbrecht2015joint]. These genomic annotations are obtained using chromatin immunoprecipitation followed by high-throughput DNA sequencing (ChIP-seq). The resolution of ChIP-seq experiments is typically on the order of tens to hundreds of bases [@mendoza2009genome], well below the resolution of Hi-C data (tens of kilobases; 750 bp is the highest resolution of Hi-C data to date [@bonev2017multiscale]). Therefore, leveraging precisely mapped genomic annotations in a supervised machine learning framework provides the possibility of more precisely predicting the locations of domain boundaries.

We present *preciseTAD*, an R package implementing an optimally tuned machine learning framework for precise identification of domain boundaries using genome annotation data. Our method utilizes the random forest (RF) algorithm trained on chromatin segmentation (BroadHMM), histone modifications (HM), and transcription factor binding sites (TFBS) data. We introduce a systematic pipeline for building optimal boundary prediction classifier. We found that spatial associations (distance) between boundaries and annotations perform best, transcription factor binding sites further improve prediction accuracy, and a simple random undersampling technique effectively addresses the negative effect of class imbalance. Translated from Hi-C data resolution level to base level (annotating each base and predicting its boundary probability), *preciseTAD* employs density-based clustering (DBSCAN) and scalable partitioning around medoids (CLARA) on prediction probabilities to select genome annotation-guided boundary coordinates. This approach circumvents resolution restrictions of Hi-C data, allowing for the precise detection of biologically meaningful boundaries. We demonstrate that *preciseTAD* predictions are more enriched for known molecular drivers of 3D chromatin, including CTCF, RAD21, SMC3, and ZNF143. Further, we show that *preciseTAD* boundaries are more conserved across cell lines, and the model trained in one cell line can predict boundaries in another cell line. This improved precision in the domain boundary location can provide insight into the association between genomic regulators and the 3D genome organization. The *preciseTAD* R package is freely available on GitHub (https://github.com/dozmorovlab/preciseTAD) and on Bioconductor (submitted).

@@

We present *preciseTAD*, a machine learning approach for precise prediction of TAD or loop boundaries from functional genomic annotations. *preciseTAD* leverages a random forest (RF) classification model built on ground truth TAD/loop boundaries obtained from domain calling tools, and genomic annotations as the predictor space. *preciseTAD* was benchmarked against two boundary calling tools, Arrowhead [@durand2016juicebox], an established TAD-caller, and Peakachu [@salameh2019supervised] a recently published algorithm for predicting chromatin loops. We first optimized our RF model by systematically comparing different combinations of genome binning (resolution), feature engineering procedures, and resampling techniques. Our results demonstrate that genomic distance between high-resolution boundary regions and transcription factor binding sites, coupled with random under-sampling results in best model performance. We show that binding of four transcription factors (SMC3, RAD21, CTCF, ZNF143) is sufficient for accurate boundary predictions. Applied to each annotated base, the model predicts the probability for each base to be a boundary. Further, coupled with spatial clustering (DBSCAN) and scalable partition around medoids clustering (CLARA), it allows for the detection of boundaries at the base-level resolution. *preciseTAD* is implemented as an R package, and allows for precise prediction of boundaries for any cell type for which the corresponding genomic annotations are available.

@@

A limitation of *preciseTAD* is that it is dependent on "ground truth" boundaries provided by a user-defined domain caller. Given the wide variety of domain callers and their variable performance [@Zufferey:2018aa; @forcato2017comparison], it is hard to define "ground truth" boundaries. <!--Ideally, we would benchmark *preciseTAD* against simulated boundaries. While methods for simulating Hi-C data sets with boundary annotations exist [@forcato2017comparison; @zheng2020freehi], methods for simulating the associated genomic annotations (the main component guiding *preciseTAD* predictions) are lacking.-->Ultra-deep Hi-C sequencing [@bonev2017multiscale] and newer technologies for precise mapping of chromatin interactions (e.g., Micro-C [@Krietenstein:2020aa]), coupled with more precise technologies for genomic annotation profiling (e.g., CUT&RUN for precise mapping of transcription factor binding sites) will help to refine the location and the genomic signatures of "ground truth" TAD boundaries. In the current work, given *preciseTAD* can be trained on the whole-genome data, we feel that the total number of TAD/loop boundaries is sufficient to guide learning of the association between genomic annotations and boundaries for precise boundary predictions.

Another limitation of *preciseTAD* is that it does not distinguish boundary types. The hierarchical nature of TAD boundaries [@weinreb2016identification; @fraser2015hierarchical; @Cresswell:2020ab] is not considered by *preciseTAD* due to the lack of gold standard of TAD hierarchy. Recent research distinguishes CTCF-associated boundaries, CTCF-negative YY1-enriched boundaries, CTCF- and YY1- depleted promoter boundaries, and the fourth class of weak boundaries largely depleted of all three features [@Krietenstein:2020aa]. Furthermore, actively transcribed regions can serve as TAD boundaries themselves, independently of CTCF binding [@Harrold:2020aa]. This may lead to some TAD boundaries being undetected by *preciseTAD* despite being detected by domain callers. Our future work will include defining separate models trained on different TAD boundary types and predicting precise locations for each of them.

>>>>

**Results.** Spatial relationship (distance in the linear genome) between boundaries and annotations has the most predictive power. Transcription factor binding sites outperformed other genome annotation types. Random under-sampling significantly improved model performance. Boundaries predicted by *preciseTAD* were more enriched for CTCF, RAD21, SMC3, and ZNF143 signal and more conserved across cell lines, highlighting their higher biological significance. The model trained in one cell line performs well in predicting boundaries in another cell line.

@@

The advent of chromosome conformation capture (3C) sequencing technologies, and its successor Hi-C, have revealed a hierarchy of the 3-dimensional (3D) structure of the human genome such as chromatin loops [@rao20143d], Topologically Associated Domains (TADs) [@nora2012spatial;@dixon2012topological], and A/B compartments [@lieberman2009comprehensive]. At the kilobase scale, chromatin loops connect gene promoters with distal enhancers, thereby regulating gene expression [@Lupianez:2015aa; @Franke:2016aa]. At the megabase scale, TADs represent regions on the linear genome, that are highly self-interacting. <!--Emerging evidence has linked TADs to critical roles in cell dynamics and cell differentiation. Studies have shown that TADs themselves are highly conserved across species and cell lines [@dixon2012topological; @krefting2018evolutionary; @nora2012spatial; @dixon2015chromatin; @rao20143d; @Schmitt2016aa].--> Disruption of boundaries demarcating TADs and loops promotes cancer [@hnisz2016activation; @taberlay2016three] and other disorders [@lupianez2016breaking;@Sun:2018aa;@Meaburn:2007aa]. Identifying the precise location of TAD and chromatin loop boundaries remains a top priority in our goal to fully understand the functionality of the human genome.

Several methods have been proposed to identify genomic coordinates of TADs (reviewed in [@Zufferey:2018aa; @forcato2017comparison]), and chromatin loops [@rao20143d; @Ay:2014aa; @Heinz:2010aa; @salameh2019supervised]. However, a key limitation among most of these TAD- and loop callers is that they depend on Hi-C data *resolution*. Resolution refers to the size of genomic regions (bins) used to segment the linear genome [@li2019comparison]. Lower resolution contribute to larger bin size, which leads to increased uncertainty in the location of boundaries. Due to high sequencing costs, most available Hi-C datasets have relatively low resolution, ranging from 25 kb to 100 kb. While these low-resolution Hi-C datasets can be used to probe higher-order genomic patterns such as A/B compartments, more refined resolutions are necessary to accurately define TAD and loop boundaries [@fudenberg2016formation]. Because increasing the resolution of Hi-C data requires a quadratic increase in the total sequencing depth [@schmitt2016genome], obtaining high resolution remains difficult.

Another limitation among domain callers is that they disregard prior knowledge about functional genomic annotations associated with TAD and loop boundaries, referred hereafter as domain boundaries. The insulator binding protein, CTCF, and additional cofactors such as SMC3 and RAD21 have been identified as components of the loop extrusion model<!--, whereby DNA is extruded through cohesin rings forming chromatin loops--> [@hansen2018recent;@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self; @Mirny:2019aa; @Davidson:2019aa]. Furthermore, distinct patterns of histone modifications have also been shown to be present at boundaries<!--including H3K36me3--> [@lieberman2009comprehensive;@dixon2012topological;@libbrecht2015joint]. These genomic annotations are obtained using chromatin immunoprecipitation followed by high-throughput DNA sequencing (ChIP-seq). The resolution of ChIP-seq experiments is typically on the order of tens to hundreds of bases [@mendoza2009genome], well below the resolution of Hi-C data (tens of kilobases; 750 bp is the highest resolution of Hi-C data to date [@bonev2017multiscale]). Therefore, leveraging precisely mapped genomic annotations in a supervised machine learning framework provides the possibility of more precisely predicting the locations of domain boundaries.

We present *preciseTAD*, an optimally tuned machine learning framework for precise identification of domain boundaries using genome annotation data. Our method utilizes the random forest (RF) algorithm trained on high-resolution chromatin segmentation (BroadHMM), histone modifications (HM), and transcription factor binding sites (TFBS) data to predict low-resolution boundaries. We introduce a systematic pipeline for building the optimal boundary prediction classifier. We found that spatial associations (linear distance) between boundaries and annotations perform best, transcription factor binding sites further improve prediction accuracy, and a simple random undersampling technique effectively addresses the negative effect of class imbalance. Translated from Hi-C data resolution level to base level (annotating each base and predicting its boundary probability), *preciseTAD* employs density-based clustering (DBSCAN) and scalable partitioning around medoids (CLARA) to select genome annotation-guided boundary coordinates. This approach circumvents resolution restrictions of Hi-C data, allowing for the precise detection of biologically meaningful boundaries. We demonstrate that *preciseTAD* predictions are more enriched for known molecular drivers of 3D chromatin, including CTCF, RAD21, SMC3, and ZNF143. Further, we show that *preciseTAD* boundaries are more conserved across cell lines, and the model trained in one cell line can predict boundaries in another cell line. This improved precision in the domain boundary location can provide insight into the association between genomic regulators and the 3D genome organization. The *preciseTAD* R package is freely available on GitHub (https://github.com/dozmorovlab/preciseTAD) and on Bioconductor (submitted).

@@

We present *preciseTAD*, a machine learning approach for precise prediction of TAD or loop boundaries from functional genomic annotations. *preciseTAD* leverages a random forest (RF) classification model built on low-resolution TAD/loop boundaries obtained from domain calling tools, and high-resolution genomic annotations as the predictor space. *preciseTAD* predicts the probability of each base being a boundary, and identifies the precise location of boundary regions and the most likely boundary points. *preciseTAD* was benchmarked against two boundary calling tools, Arrowhead [@durand2016juicebox], an established TAD-caller, and Peakachu [@salameh2019supervised] a recently published algorithm for predicting chromatin loops. We first optimized our RF model by systematically comparing different combinations of genome binning (resolution), feature engineering procedures, and resampling techniques. Our results demonstrate that genomic distance between high-resolution boundary regions and transcription factor binding sites, coupled with random under-sampling results in best model performance. We show that binding of four transcription factors (SMC3, RAD21, CTCF, ZNF143) is sufficient for accurate boundary predictions. *preciseTAD* is implemented as an R package, and allows for precise prediction of boundaries for any cell type for which the corresponding genomic annotations are available.

@@

A limitation of *preciseTAD* is that it is dependent on "ground truth" boundaries provided by a user-defined domain caller. Given the wide variety of domain callers and their variable performance [@Zufferey:2018aa; @forcato2017comparison], it is hard to define "ground truth" boundaries. <!--Ideally, we would benchmark *preciseTAD* against simulated boundaries. While methods for simulating Hi-C data sets with boundary annotations exist [@forcato2017comparison; @zheng2020freehi], methods for simulating the associated genomic annotations (the main component guiding *preciseTAD* predictions) are lacking.-->Ultra-deep Hi-C sequencing [@bonev2017multiscale] and newer technologies for precise mapping of chromatin interactions (e.g., Micro-C [@Krietenstein:2020aa]), coupled with more precise technologies for genomic annotation profiling (e.g., CUT&RUN for precise mapping of transcription factor binding sites) will help to refine the location and the genomic signatures of "ground truth" boundaries. In the current work, given *preciseTAD* can be trained on the whole-genome data, we feel that the total number of TAD/loop boundaries is sufficient to guide learning of the association between genomic annotations and boundaries for precise boundary predictions.

Another limitation of *preciseTAD* is that it does not distinguish boundary types. The hierarchical nature of TAD boundaries [@weinreb2016identification; @fraser2015hierarchical; @Cresswell:2020ab] is not considered by *preciseTAD* due to the lack of gold standard of TAD hierarchy. *preciseTAD* also does not consider the directionality of CTCF binding [@Harrold:2020aa] as it predicts individual boundaries in contrast to pairs of convergent CTCF motifs marking individual domains. Recent research distinguishes CTCF-associated boundaries, CTCF-negative YY1-enriched boundaries, CTCF- and YY1- depleted promoter boundaries, and the fourth class of weak boundaries largely depleted of all three features [@Krietenstein:2020aa]. Furthermore, actively transcribed regions can serve as TAD boundaries themselves, independently of CTCF binding [@Harrold:2020aa]. This may lead to some TAD boundaries being undetected by *preciseTAD* despite being detected by domain callers. Our future work will involve incorporating the directionality of CTCF binding in predictive modeling, including additional predictor types, and defining separate models trained on different boundary types.
